Human computational hysteresis: measuring inter-task effects

### 2014-09-23 docs/publications/2014.09_SCIENCE/man.tex
backup work on manuscript

\usepackage{graphicx}
\usepackage{amsthm}
\newtheorem*{mydef}{Definition}
\usepackage{dcolumn}
\usepackage{multirow}
\usepackage{booktabs}
\newcolumntype{d}{D{.}{.}{4.0}}
\newcolumntype{s}{D{.}{.}{1.4}}

@@

\begin{sciabstract}
Microtask platforms offer a fast, flexble, and inexpensive source of labour for
clerical tasks \cite{Finnerty2013, chandler2013breaking, Berinsky2012351} and 
recently have been adopted by researchers to augment datasets and solicit 
participants \cite{paolacci2010running, Berinsky2012351, chandler2013breaking}.
Some computer scientists describe such platforms as a new computing 
architecture, and seek to characterize “human co-processing units” (HPUs) in 
analogy to CPUs and GPUs \cite{5543192}. However, it is known that people are 
susceptible to 
priming effects\cite{No2007,Swaab200299,Gopher2000308,sohn2001task,Ghuman17062008,BJOP1796,BJOP1826,Gass1999549}, which stands to complicate the HPU 
perspective.  Here, we introduce a framework to recast this psychological 
phenomenon as a kind of computational hysteresis. 
We apply this framework to study \textit{inter-task effects} that arise when 
microtask 
workers perform a sequence of tasks. Surprisingly, earlier tasks can have a 
very strong influence on workers during later tasks---stronger than overtly
framing the tasks as serving different research goals.   
This result has major implications for the design of crowdsourcing initiatives.
The framework we introduce provides a general measure of priming severity, 
and, in combination with machine learning tools, enables measurement without
knowing ahead of time how priming will affect output.
\end{sciabstract}

@@

\paragraph*{Priming as computational hysteresis.}
In psychology, priming effects are studied because they hint at the 
psychological mechanisms at play during certain cognitive tasks. But in the 
context of human computation, the focus should be on the outward consequences: 
the potential for algorithms to execute differently when ``run'' on differently
primed HPUs.

We will present three definitions of priming with this practical concern in 
mind, and then describe how to measure priming in practice. To proceed with 
the definitions, we will assume that two populations of HPU’s, $\mathcal{P}$ 
and $\mathcal{Q}$ were 
sampled randomly from the same initial population $\mathcal{R}$, and then were 
exposed to (possibly) different priming treatments. The HPUs in $\mathcal{P}$ 
and $\mathcal{Q}$ perform some 
task, leading to popualtions of work-products, $P$ and $Q$.

For the first defenition, let us address the concern that differently primed 
HPUs may produce ``biased'' work-products.

\begin{mydef}
	{\upshape Maximum Bias $\theta_\mathrm{bias}$:}
	Let $\mathcal{D}$ be a binary decision algorithm, which takes a 
	work-product
	$r \in P \cup Q$ as input, and outputs a bit $\mathcal{D}(r)=b$, where 
	$b \in \{0,1\}$. Denote the \emph{bias} of $\mathcal{D}$, due to the
	different priming of $\mathcal{P}$ and $\mathcal{Q}$ by
	$\theta^\mathcal{D}_\mathrm{bias}(P,Q)$ and define it to be the difference 
	in the expected output of $\mathcal{D}$ when acting on work-products from 
	$P$ vs $Q$:
	$$
	\theta^\mathcal{D}_\mathrm{bias}(P,Q) = 
		\left| 
			\mathrm{E}\left\{ \mathcal{D}(p) \right\} 
			- \mathrm{E}\left\{ \mathcal{D}(q) \right\} 
		\right|,
	$$
	where $p \in P$, $q \in Q$.  Then, define the difference in priming 
	between $P$ and $Q$ to be the worst-case bias, i.e. the bias of the most
	susceptible decision:
	$$
	\theta_\mathrm{bias} = 
		\sup_\mathcal{D} \theta^\mathcal{D}_\mathrm{bias}(P,Q)
	$$
\end{mydef}

Thus, $\theta_\mathrm{bias}$ has a straightforward interpretation as an upper 
bound on the amount of bias induced on any yes-or-no decision, due to the 
difference in priming of $\mathcal{P}$ and $\mathcal{Q}$.

For the next definition, we attempt to address directly the extent to which 
two populations of work-products are ``different'', by supposing that they are 
different to the extent that they can be distinguished algorithmically. To 
facilitate this definition, we make use of the notions of a 
\textit{distinguisher} and a \textit{validation} test.

\begin{mydef}
	{\upshape Algorithmic distinctness $\theta_\mathrm{dist}$:}
	Let $\mathcal{D}$ be a distinguisher algorithm, which takes work-product 
	$r \in P \cup Q$ as input, and outputs a \emph{guess} 
	$\mathcal{D}(r) \in \{0, 1\}$. Now define a validation test 
	$V(\mathcal{D}, R_0, R_1)$ as follows. A bit is drawn uniformly at random, 
	$b \in_R \{0, 1\}$, and then a work-product $r$ is chosen uniformly at 
	random from one of the populations, decided by the bit: $r \in_R R_b$. 
	Feed $r$ to $\mathcal{D}$, and if $\mathcal{D}(r) = b$, we say 
	$\mathcal{D}$ guessed correctly, and the value of $V$ is 1, 
	otherwise V = 0.

	A distinguisher that guesses randomly will still be correct half of the 
	time. Therefore, denote the performance of $\mathcal{D}$ as 
	$\eta_\mathcal{D}(P, Q)$, and define:
	$$ 
		\theta_\mathrm{dist} = \sup_\mathcal{D} \eta_\mathcal{D}(P,Q)
	$$

\end{mydef}
In the above definition, the performance simply rescales accuracy, so that when
accuracy is 1/2 (i.e. as good as chance) performance is 0, and when accuracy 
is perfect, performance is 1. This definition casts the measurement of priming 
effects in terms of a binary classification problem, which is well-studied in 
the field of machine learning.

The two definitions we have already mentioned are, in fact, equivalent, and 
correspond to a third, more fundamental formulation known as the total 
\textit{variational distance}, or $l_1$-distance, or simply the “statistical 
distance” ($D_{l_1}$):
$$
	D_{l_1}(P,Q) \equiv \frac{1}{2} \sum_{x \in \mathcal{X}} 
	\left| 
		f_P(x) - f_Q(x)
	\right|,
$$
where $f_P(x)$ denotes the probability of observing x when sampling from P. 
This leads us to our third definition:

\begin{mydef}
	{\upshape Priming as $l_1$-distance:}
	The difference in priming of $\mathcal{P}$ and $\mathcal{Q}$ is the 
	$l_1$-distance between the distributions of their work-products, 
	$D_{l_1}(P,Q)$.
\end{mydef}

We provide proofs of the equivalence between these definitions in the 
supplementary material. Despite their equivalence, we show the three 
definitions because they offer different theoretical, practical, and 
interpretive entry points. Definition 1 provides a practical motivation, and 
the straighforward interpretation of priming as bias. Definition 3 draws a 
clear link to existing statistical theory. And Definition 2 introduces machine 
learning tools, which allow us to gain traction when the purely statistical 
tools fail.

\paragraph{Measuring priming.}
Care must be taken in estimating $D_{l_1}$ from samples so as not to obtain
an inflated estimate. In the so-called naive approach, one uses the frequency 
with which a given work-product $x$ is obtained from HPUs in $\mathcal{P}$ to 
estimate the probability mass function at $x$; 
$\hat{f}_P(x) = (\sum_{X\in P} \mathbf{1}_{[X=x]})/|P|$.
One can then estimate 
$D_{l_1} = \sum_{x \in \mathcal{X}} \left|\hat{f}_P(x) - \hat{f}_Q(x) \right|/2$ using only the sample frequencies. But the number of samples, $|P|$ and 
$|Q|$, must be much larger than the number of possible work-outputs, 
$|\mathcal{X}|$, otherwise, this method produces an enormously inflated 
estimate.

>>>>

\usepackage{graphicx}
\usepackage{amsthm}
\newtheorem{mydef}{Definition}
\usepackage{dcolumn}
\usepackage{multirow}
\usepackage{booktabs}
\newcolumntype{d}{D{.}{.}{4.0}}
\newcolumntype{s}{D{.}{.}{1.4}}

@@

\begin{sciabstract}
Microtask platforms offer a fast, flexble, and inexpensive source of labour for
clerical tasks \cite{Finnerty2013, chandler2013breaking, Berinsky2012351} and 
recently have been adopted by researchers to augment datasets and solicit 
participants \cite{paolacci2010running, Berinsky2012351, chandler2013breaking}.
Computer scientists have described such platforms as a new computing 
architecture, and seek to characterize “human co-processing units” (HPUs) in 
analogy to CPUs and GPUs \cite{5543192}. However, it is known that people are 
susceptible to 
priming effects\cite{No2007,Swaab200299,Gopher2000308,sohn2001task,Ghuman17062008,BJOP1796,BJOP1826,Gass1999549}, which stands to complicate the HPU 
perspective.  Here, we introduce a framework to recast this psychological 
phenomenon as a kind of computational hysteresis. 
We apply this framework to study \textit{inter-task effects} that arise when 
microtask 
workers perform a sequence of tasks. Surprisingly, earlier tasks can have a 
very strong influence on workers during later tasks---stronger than overtly
framing the tasks as serving different research goals.   
This result has major implications for the design of crowdsourcing initiatives.
The framework we introduce provides a general measure of priming severity, 
and, in combination with machine learning tools, enables measurement without
knowing ahead of time how priming will affect output.
\end{sciabstract}

@@

\paragraph*{Priming as computational hysteresis.}
To begin, we adopt the perspective that there is no canonical neutral or
unprimed state.  Rather, we will speak of one HPU population being 
\textit{differently primed} from another.  We present three definitions for 
the difference in priming between HPU populations, each having different 
merits, and then we discuss their relationships. To proceed with the 
definitions, we will assume that two populations of HPU’s, $\mathcal{P}$ and 
$\mathcal{Q}$ were 
sampled randomly from the same initial population $\mathcal{R}$, and then were 
exposed to (possibly) different priming treatments. The HPUs in $\mathcal{P}$ 
and $\mathcal{Q}$ perform some task, leading to popualtions of work-products, 
$P$ and $Q$.

For the first defenition, we focus on the direct effects that running an 
algorithm on differently primed HPUs has on the algorithm's output.

\begin{mydef}
	\label{def:bias}
	{\upshape Maximum Bias $\theta_\mathrm{bias}$:}
	Let $\mathcal{D}$ be a binary decision algorithm, which takes a 
	work-product
	$r \in P \cup Q$ as input, and outputs a bit $\mathcal{D}(r)=b$, where 
	$b \in \{0,1\}$. Denote the \emph{bias} of $\mathcal{D}$, due to the
	different priming of $\mathcal{P}$ and $\mathcal{Q}$ by
	$\theta^\mathcal{D}_\mathrm{bias}(P,Q)$ and define it to be the difference 
	in the expected output of $\mathcal{D}$ when acting on work-products from 
	$P$ and $Q$:
	$$
	\theta^\mathcal{D}_\mathrm{bias}(P,Q) = 
		\left| 
			\mathrm{E}\left\{ \mathcal{D}(p) \right\} 
			- \mathrm{E}\left\{ \mathcal{D}(q) \right\} 
		\right|,
	$$
	where $p \in P$, $q \in Q$.  Then, define the difference in priming 
	between $\mathcal{P}$ and $\mathcal{Q}$ to be the worst-case bias, i.e. 
	the bias of the most susceptible decision:
	$$
	\theta_\mathrm{bias} = 
		\sup_\mathcal{D} \theta^\mathcal{D}_\mathrm{bias}(P,Q)
	$$
\end{mydef}
Thus, in \textbf{Definition \ref{def:bias}}, $\theta_\mathrm{bias}$ has a 
straightforward interpretation as an upper bound on the amount of bias induced 
on any yes-or-no decision, due to the difference in priming of 
$\mathcal{P}$ and $\mathcal{Q}$.  Although easy to interpret, it is, however, 
not clear how $\theta_\mathbf{bias}$ could be measured in practice.

Next we attempt to phase the difference in priming between $\mathcal{P}$ and
$\mathcal{Q}$ in terms of the extent to which their work products are 
distinguishable algorithmically. To facilitate this definition, we make 
use of the notions of a \textit{distinguisher} and a \textit{validation} test.
\begin{mydef}
	{\upshape Algorithmic distinctness $\theta_\mathrm{dist}$:}
	Let $\mathcal{D}$ be a distinguisher algorithm, which takes work-product 
	$r \in P \cup Q$ as input, and outputs a \emph{guess} 
	$\mathcal{D}(r) \in \{0, 1\}$. Now define a validation test 
	$V(\mathcal{D}, R_0, R_1)$ as follows. A bit is drawn uniformly at random, 
	$b \in_R \{0, 1\}$, and then a work-product $r$ is chosen uniformly at 
	random from one of the populations, decided by the bit: $r \in_R R_b$. 
	Feed $r$ to $\mathcal{D}$, and if $\mathcal{D}(r) = b$, we say 
	$\mathcal{D}$ guessed correctly, and the value of $V$ is 1, 
	otherwise $V = 0$.

	A distinguisher that guesses randomly will still be correct half of the 
	time. Therefore, denote the performance of $\mathcal{D}$ as 
	$\eta_\mathcal{D}(P, Q)$:
	$$
		\eta_\mathcal{D}(P,Q) = 2\cdot \mathrm{E}\{V(\mathcal{D},P,Q)\} - 1
	$$
	Then define the difference in priming between $\mathcal{P}$ and 
	$\mathcal{Q}$ to be the performance of the best possible distinguisher:
	$$ 
		\theta_\mathrm{dist} = \sup_\mathcal{D} \eta_\mathcal{D}(P,Q)
	$$
\end{mydef}
In the above definition, the performance simply rescales accuracy, so that when
accuracy is 1/2 (i.e. as good as chance) performance is 0, and when accuracy 
is perfect, performance is 1. This definition formulates the measurement of 
priming effects in terms of a binary classification problem. As we will show,
this allows us to apply tools from the machine learning field to the 
measurement of difference in priming.  However, it is not immediately clear 
how this measure relates to the potential for priming to cause bias.

For the third definition, we focus on describing the inherent, statistical
difference between work-products in $P$ and $Q$. Thus we define the difference
in priming to be equal to the \textit{total variational distance}, or $l_1$-distance (denoted $D_{l_1}$) between $P$ and $Q$. 
\begin{mydef}
	\label{def:l1}
	{\upshape total variational distance $\theta_{l_1}$:}
	$$
	\theta_{l_1} = D_{l_1}(P,Q) \equiv \frac{1}{2} \sum_{x \in \mathcal{X}} 
	\left| 
		f_P(x) - f_Q(x)
	\right|,
	$$
	where $f_P(x)$ denotes the probability of observing x when sampling from P.
\end{mydef}
The abvantage of definition \ref{def:l1} is that it provides connections
to statistical and information theory.

Remarkably, all three definitions are, in fact, equivalent, and from here on
we simply denote the difference in priming as $\theta$.  We provide proofs
in the supplementary material to demostrate this fact.  This means that we
can use machine learning tools to measure priming, while interpreting the
results either in terms of the  leakage of information from a prime into an 
HPU output or in terms of the potential to bias a decision.

\paragraph{Measuring priming.}
Before proceeding, we offer a few words of caution regarding the measurement 
of $\theta$.

Care must be taken in estimating $D_{l_1}$ from samples so as not to obtain
an inflated estimate. In the so-called naive approach, one uses the frequency 
with which a given work-product $x$ is obtained from HPUs in $\mathcal{P}$ to 
estimate the probability mass function at $x$; 
$\hat{f}_P(x) = (\sum_{X\in P} \mathbf{1}_{[X=x]})/|P|$.
One can then estimate 
$D_{l_1} = \sum_{x \in \mathcal{X}} \left|\hat{f}_P(x) - \hat{f}_Q(x) \right|/2$ using only the sample frequencies. But the number of samples, $|P|$ and 
$|Q|$, must be much larger than the number of possible work-outputs, 
$|\mathcal{X}|$, otherwise, this method produces an enormously inflated 
estimate.
